---
title: "Assignment 2: Coding Basics"
author: "Hanna Karnei"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
# Create a sequence of numbers from 1 to 30, increasing by 3
my.sequence <- seq (1,30,3)

# Calculate mean and median of the sequence
mean(my.sequence)
median(my.sequence)

# Check if the mean is greater than the median
mean(my.sequence) > median(my.sequence)

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
# Create a series of vectors
names <- c('Hanna', "Haru", "Jolina", "John", "Juan") 
test.scores <- c(100, 90, 45, 85, 40) 
pass <- c(TRUE, TRUE, FALSE, TRUE, FALSE) 

# Check vector type
class(names) #character vector
class(test.scores) #numeric vector
class(pass) #logical vector

# Combine vectors into a dataframe and assign names to columns 
student.test.info <- data.frame(Name=names, Score=test.scores, Pass=pass)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: The main difference between a matrix and a dataframe is that a dataframe can stroce different types of vector (e.g. logical, numeric, character), whereas a matrix stores only one type of data. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

my.function <- function(x){
  ifelse(x<50, FALSE, TRUE)
}

check.for.pass <- my.function(test.scores)

```


12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: "ifelse" worked because this statement can handle vectors. In contrast, the "if" and "else" statement can only be applied to a single value, so it would not work in our case because we have a list of test scores we want to test for the True/False condition. 
