---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Hanna Karnei"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1 Load libraries 
getwd()
library(tidyverse)
library (agricolae)
library(lubridate)
library(here)
library(dplyr)
here()

lake.chem.physics <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
                              stringsAsFactors = TRUE)

head(lake.chem.physics)

lake.chem.physics$sampledate <- as.Date(lake.chem.physics$sampledate, format = "%m/%d/%y")

#2.Build a theme, set it as default

mytheme <- theme_classic(base_size = 12) +
  theme(axis.text = element_text(color = "grey"), 
        legend.position = "top")

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:
H0: Lake temperature recorded in July does not change with depth across all lakes 

>Ha: Lake temperature recorded in July changes with depth across all lakes 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 Filter dataframe
filtered.lake <- lake.chem.physics %>%
  filter(month(sampledate) == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit

head(filtered.lake)

#5 Plot a scatterplot

plot.depth.temp<-
  ggplot(filtered.lake, aes(x = depth, y = temperature_C)) +
  geom_smooth(method = "lm") +
  ylim(0, 35) +
  ggtitle('Relationship Between Depth and Temperature') + 
  ylab('Temperature (C)') + 
  xlab ('Depth')+
  geom_point() 

print(plot.depth.temp) 

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The plot suggests that there is a negative correlation between depth and temeperature. As depth increases, temperature declines. The distribution of points suggests that the trend is not perfectly linear, as data points are scattered unevenly above and below the trend line. 

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7 Linera regression
depth_temp_regression <- lm(data = filtered.lake, temperature_C ~ depth)
summary(depth_temp_regression)
```

8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model results suggest that 73.9% of variability in temperature can be attrubted to changes in depth. The degrees of freedom for this test is 9726. The result is statistically significant (p-value is < 0.001). Temeprature is predicted to change by 1.946 degrees celcius for every 1 meter change in depth. 
 
---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 AIC model
AIC.model <- lm(data = filtered.lake, temperature_C ~ year4 + daynum + depth)
step(AIC.model)

#10 Multiple linear regression model
final.model <- lm(data = filtered.lake, temperature_C ~ year4 + daynum + depth)
summary(final.model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables suggested by the AIC model are all initial variables: year4, daynum and depth. AIC and RRS values are the lowest when no variable is excluded from the model. This model explains 74.1% of the variability in temperature, which is only a slight imrovement from the simple regression model with depth as the only explanatory variable.  

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12 Run ANOVA

lakes.anova <- aov(data = filtered.lake, temperature_C ~ lakename)
summary(lakes.anova)

lakes.anova2 <- lm(data = filtered.lake, temperature_C ~ lakename)
summary(lakes.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, the ANOVA test suggests that there is a signficant difference in mean temperature among the lakes. This is indicated by a p-value less than 0.001. The linear model provides details for the difference in mean temperature by lake, and again we can see that the diffrences between all lakes are statistically significant. 

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.height = 6, fig.width = 9}
#14. Plot the correlation by lake

plot.depth.temp.by.lake<-
  ggplot(filtered.lake, aes(x = depth, y = temperature_C, color=lakename)) +
  geom_smooth(method = "lm", se=FALSE) +
  ylim(0, 35) +
  ggtitle('Association Between Depth and Temperature by Lake') + 
  ylab('Temperature (C)') + 
  xlab ('Depth')+
  geom_point(aes(color = lakename), alpha = 0.5) +
  geom_point() 

print(plot.depth.temp.by.lake) 

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15 Run Tukey HSD test

TukeyHSD(lakes.anova)

lakes.groups <- HSD.test(lakes.anova, "lakename", group = TRUE)
lakes.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward Lake and Paul Lake have the same mean temperature, statistically speaking. No lake has a mean that is statistically different from other lakes. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could use an independent two sample T-test to compare the mean temperatures in the lakes. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#18. Perform two sample t-test 

filtered.lake.2 <- filtered.lake %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

ttest <- t.test(filtered.lake.2$temperature_C ~ filtered.lake.2$lakename)
ttest
```

>Answer: The results of the t-test show that the average July temperatures between two lakes are not statistically different. The p-value of 0.26 suggest that the null hypothesis cannot be rejected. This result confirms the findings of Tukey's HSD test -- both lakes had an overlapping letter "b" (Crampton lake - "ab", Ward Lake - "bc"), indicating that their mean temperature are the same, statistically speaking.
